Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-03-14 Paper/main.tex
Major changes to the text and structure of the paper. Added TODO.txt

\usepackage{natbib}
\bibliographystyle{agsm}
\let\hardvardurl\url
\usepackage{url}
\usepackage[breaklinks]{hyperref}

@@

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed. In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=14cm]{pictures/screenshot-gtranslate-hungarian}}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
	\label{fig:screenshot-gtranslate-hungarian}
\end{figure}

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure. The researchers are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

@@

We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}). Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) are shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset. Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.

\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Korean 				& O 				& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
									& Ido 					& O 				& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
	\end{small}
\end{table}

\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|m{3.5cm}|c|c|c|}
	\hline	
	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the U.S. Bureau of Labor Statistics \url{https://???}, grouped by category. We obtained a total of $1019$ occupations from $22$ distinct categories. We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
	\label{tab:occupations}
	\end{small}
\end{table}

\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Category 		& \# Occupations 	\\ \hline \hline
	Artistic 		& $102$ 			\\ \hline
	Computer 		& $19$ 				\\ \hline
	Corporate 		& $50$ 				\\ \hline
	Dance 			& $9$ 				\\ \hline
	Film/Television & $26$ 				\\ \hline
	Healthcare 		& $88$ 				\\ \hline
	Industrial 		& $26$ 				\\ \hline
	Science 		& $50$ 				\\ \hline
	Service 		& $10$ 				\\ \hline
	Theatre 		& $52$ 				\\ \hline
	Writing 		& $29$ 				\\ \hline
	\hline
	Total			& $462$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category. We have selected a total of 462 occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
	\label{tab:occupations}
	\end{small}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Language 	& Sentence template \\ \hline \hline
	Malay		& dia adalah $\langle occupation \rangle$ \\ \hline
	Estonian	& ta on $\langle occupation \rangle$ \\ \hline
	Finnish		& hän on $\langle occupation \rangle$ \\ \hline
	Hungarian	& ő $\langle occupation \rangle$ \\ \hline
	Armenian	& na $\langle occupation \rangle$ e \\ \hline
	Japanese	& $\langle occupation \rangle$ です\\ \hline
	Turkish		& o bir $\langle occupation \rangle$ \\ \hline
	Yoruba		& o jẹ $\langle occupation \rangle$ \\ \hline
	Basque		& $\langle occupation \rangle$ da \\ \hline
	Swahili		& yeye ni $\langle occupation \rangle$ \\ \hline
	Chinese		& ta $\langle occupation \rangle$ \\ \hline
	\end{tabular}
	\caption{Templates used to infer gender biases in the translation to the English language.}
	\label{tab:templates}
	}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|}
	\hline
	Insurance sales agent & Editor & Rancher \\ \hline
	Ticket taker & Pile-driver operator & Tool maker \\ \hline
	Jeweler & Judicial law clerk & Auditing clerk \\ \hline
	Physician & Embalmer & Door-to-door salesperson \\ \hline
	Packer & Bookkeeping clerk & Community health worker \\ \hline
	Sales worker & Floor finisher & Social science technician \\ \hline
	Probation officer & Paper goods machine setter & Heating installer \\ \hline
	Animal breeder & Instructor & Teacher assistant \\ \hline
	Statistical assistant & Shipping clerk & Trapper \\ \hline
	Pharmacy aide & Sewing machine operator & Service unit operator \\ \hline
	\end{tabular}
	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
	\label{tab:occupations-examples}
	}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|}
	\hline
	happy 		& sad 		& right 	\\ \hline
	wrong 		& afraid	& brave 	\\ \hline
	smart		& dumb		& proud 	\\ \hline
	ashamed		& kind		& cruel 	\\ \hline
	envious		& loving	& hateful 	\\ \hline
	modest 		& arrogant	& guilty	\\ \hline
	innocent	& helpless	& shy		\\ \hline
	\end{tabular}
	\caption{Adjectives}
	\label{tab:adjectives}
	}
\end{table}

\section{Results}

>>>>

\usepackage{natbib}
%\bibliographystyle{agsm}
\bibliographystyle{named}
\let\hardvardurl\url
\usepackage{url}
\usepackage[breaklinks]{hyperref}

@@

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal assymetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed. In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
	\label{fig:screenshot-gtranslate-hungarian}
\end{figure}

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique. They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

@@

% The hungarian punctuation in 'ő' is not rendered in the PDF. This needs to be fixed somehow. The same problem extends to Table 4 (label tab:templates)
We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when \say{nurse} is replaced by \say{engineer}. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work 
% PEDRO: Removed "with as many gender neutral languages as possible"
 with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures \cite{wals} and other sources. Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
(1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. This time we resorted to using the Bureau of Labor Statistics' detailed occupations table \cite{BLS2017}, from the United States Department of Labor, for providing such a list. The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones. Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers. We then separated the curated jobs into categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset. Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.

% Table containing a list of gender neutral languages with information about pronominal gender systems and GTranslate support
\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
									& English 				& \checkmark 				& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Korean 				& O 						& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
	\end{small}
\end{table}

% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|m{3.5cm}|c|c|c|}
	\hline	
	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the U.S. Bureau of Labor Statistics \url{https://???}, grouped by category. We obtained a total of $1019$ occupations from $22$ distinct categories. We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
	\label{tab:occupations}
	\end{small}
\end{table}

% List of sentence templates for assessing gender bias in the translation from source language to English
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Language 	& Sentence template 						\\ \hline \hline
	Malay		& dia adalah $\langle occupation \rangle$ 	\\ \hline
	Estonian	& ta on $\langle occupation \rangle$ 		\\ \hline
	Finnish		& hän on $\langle occupation \rangle$ 		\\ \hline
	Hungarian	& ő $\langle occupation \rangle$ 			\\ \hline
	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
	Japanese	& $\langle occupation \rangle$ です			\\ \hline
	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
	Yoruba		& o jẹ $\langle occupation \rangle$ 		\\ \hline
	Basque		& $\langle occupation \rangle$ da 			\\ \hline
	Swahili		& yeye ni $\langle occupation \rangle$ 		\\ \hline
	Chinese		& ta $\langle occupation \rangle$ 			\\ \hline
	\end{tabular}
	\caption{Templates used to infer gender biases in the translation to the English language.}
	\label{tab:templates}
	}
\end{table}

% Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|}
	\hline
	Insurance sales agent 	& Editor 						& Rancher 					\\ \hline
	Ticket taker 			& Pile-driver operator 			& Tool maker 				\\ \hline
	Jeweler 				& Judicial law clerk 			& Auditing clerk 			\\ \hline
	Physician 				& Embalmer 						& Door-to-door salesperson 	\\ \hline
	Packer 					& Bookkeeping clerk 			& Community health worker 	\\ \hline
	Sales worker 			& Floor finisher 				& Social science technician \\ \hline
	Probation officer 		& Paper goods machine setter 	& Heating installer 		\\ \hline
	Animal breeder 			& Instructor 					& Teacher assistant 		\\ \hline
	Statistical assistant 	& Shipping clerk 				& Trapper 					\\ \hline
	Pharmacy aide 			& Sewing machine operator 		& Service unit operator 	\\ \hline
	\end{tabular}
	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
	\label{tab:occupations-examples}
	}
\end{table}

% Table with all adjectives we experimented with
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|}
	\hline
	Happy 		& Sad 		& Right 	\\ \hline
	Wrong 		& Afraid	& Brave 	\\ \hline
	Smart		& Dumb		& Proud 	\\ \hline
	Ashamed		& Kind		& Cruel 	\\ \hline
	Envious		& Loving	& Hateful 	\\ \hline
	Modest 		& Arrogant	& Guilty	\\ \hline
	Innocent	& Helpless	& Shy		\\ \hline
	\end{tabular}
	\caption{Adjectives}
	\label{tab:adjectives}
	}
\end{table}

\section{Distribution of translated gender pronouns per occupation category}

A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field. What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability 
% Add statistical test here
than it does either female or gender-neutral pronouns. Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes,
% Add reference here
such as life and physical sciences, architecture, engineering, computer science and mathematics. Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation. For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).

% Table with our gender bias results grouped by category
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
	Category 								& Female ($\%$)	& Male ($\%$)	& Neutral ($\%$) \\ \hline
	\hline		
	Education / Training / Library 			& $25.207$ 		& $59.917$		& $2.893$		\\ \hline
	Business / Financial 					& $6.324$ 		& $67.194$		& $7.312$		\\ \hline
	Office / Administrative					& $10.031$		& $62.382$		& $7.732$		\\ \hline
	Healthcare support						& $29.545$		& $51.705$		& $7.955$		\\ \hline
	Management								& $9.486$		& $68.182$		& $3.953$		\\ \hline
	Installation / Maintenance / Repair		& $3.596$		& $68.332$		& $11.588$		\\ \hline
	Healthcare practitioners / technical	& $21.564$		& $61.734$		& $6.977$		\\ \hline
	Community and social service			& $15.584$ 		& $63.636$		& $4.545$		\\ \hline
	Sales and related						& $8.117$ 		& $69.156$		& $8.766$		\\ \hline
	Production								& $14.36$ 		& $53.306$		& $13.120$		\\ \hline
	Architecture and engineering			& $2.194$		& $79.624$		& $5.016$		\\ \hline
	Life, physical, and social sciences		& $3.743$ 		& $82.888$		& $3.743$		\\ \hline
	Transportation / material moving		& $7.532$ 		& $66.234$		& $11.688$		\\ \hline
	Arts / Entertainment					& $7.862$ 		& $71.007$		& $4.423$		\\ \hline
	Legal									& $11.688$ 		& $67.532$		& $5.195$		\\ \hline
	Protective service						& $6.818$ 		& $75.000$		& $5.844$		\\ \hline
	Food prepaparation / serving			& $16.578$ 		& $59.893$		& $10.160$		\\ \hline
	Farming, fishing, and forestry			& $10.490$ 		& $62.238$		& $8.392$		\\ \hline
	Computer and mathematical				& $2.841$ 		& $76.136$		& $4.545$		\\ \hline
	Personal care and service				& $20.937$ 		& $55.096$		& $9.091$		\\ \hline
	Construction and extraction				& $9.492$ 		& $61.631$		& $11.096$		\\ \hline
	Cleaning / Maintenance					& $11.818$		& $63.636$		& $8.182$		\\ \hline
	\hline
	Total           						& $11.107$ 		& $63.030$ 		& $9.278$ 		\\ \hline
	\end{tabular}
	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}. Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
	\label{tab:gender-by-category}
	}
\end{table}

% Table with our gender bias results grouped by category (but categories are further grouped into "super-categories")
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
	Category 								& Female ($\%$)			& Male ($\%$)			& Neutral ($\%$)	\\ \hline
	\hline
	Education								& $25.207$				& $59.917$				& $2.893$			\\ \hline
	Corporate								& $7.955$				& $68.030$				& $6.364$			\\ \hline
	Service									& $9.221$				& $64.987$				& $9.506$			\\ \hline
	Healthcare								& $23.729$				& $59.014$				& $7.242$			\\ \hline
	Production								& $14.360$				& $53.306$				& $13.120$			\\ \hline
	STEM									& $2.992$				& $80.322$				& $4.373$			\\ \hline
	Arts / Entertainment					& $7.862$				& $71.007$				& $4.423$			\\ \hline
	Legal									& $11.688$				& $67.532$				& $5.195$			\\ \hline
	Farming / Fishing / Forestry			& $10.490$				& $62.238$				& $8.392$			\\ \hline
	Construction / Extraction				& $9.492$				& $61.631$				& $11.096$			\\ \hline
	\hline
	Total           						& $11.107$ 				& $63.030$ 				& $9.278$ 			\\ \hline
	\end{tabular}
	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}. Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
	\label{tab:gender-by-category-grouped}
	}
\end{table}

Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed. The histogram in Figure \ref{fig:histogram-female} (and its coalesced variant in Figure \ref{fig:histogram-female-grouped}) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures \ref{fig:histogram-neutral} and \ref{fig:histogram-neutral-grouped} --, while the same data for male pronouns (shown in Figures \ref{fig:histogram-male}, \ref{fig:histogram-male-grouped}) suggests a skew normal distribution. Furthermore we can see both on Figures \ref{fig:histogram-female-grouped} and \ref{fig:histogram-male-grouped} how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.
